---
title: "CIFTI in R, MSC single-subject"
author: "Micalea Chan"
date: "3/5/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("~/gitlocal/cifti/", repos = NULL, type="source")
library(cifti)
library(tidyverse)
library(pals)
library(superheat)
library(igraph)

mscfile <- "../data/MSC/derivatives/surface_pipeline/sub-MSC01/processed_restingstate_timecourses/ses-func01/cifti/sub-MSC01_ses-func01_task-rest_bold_32k_fsLR.dtseries.nii"
tmask_file <- "../data/MSC/derivatives/surface_pipeline/sub-MSC01/processed_restingstate_timecourses/ses-func01/cifti/sub-MSC01_ses-func01_task-rest_bold_32k_fsLR_tmask.txt"

parcel_file <- "../data/MSC/derivatives/surface_pipeline/sub-MSC01/surface_parcellation/sub-MSC01_parcels.dtseries.nii"
comm_file <- "../data/MSC/derivatives/surface_pipeline/sub-MSC01/surface_parcellation/sub-MSC01_parcel_networks.dscalar.nii"

if(!file.exists(mscfile)){
  dir.create(dirname(mscfile), recursive = T)
  download.file(url = "https://openneuro.org/crn/datasets/ds000224/snapshots/00002/files/derivatives:surface_pipeline:sub-MSC01:processed_restingstate_timecourses:ses-func01:cifti:sub-MSC01_ses-func01_task-rest_bold_32k_fsLR.dtseries.nii", destfile = mscfile)
}

if(!file.exists(tmask_file)){
  dir.create(dirname(tmask_file), recursive = T)
  download.file(url = "https://openneuro.org/crn/datasets/ds000224/snapshots/00002/files/derivatives:surface_pipeline:sub-MSC01:processed_restingstate_timecourses:ses-func01:cifti:sub-MSC01_ses-func01_task-rest_bold_32k_fsLR_tmask.txt", destfile = tmask_file)
}

if(!file.exists(parcel_file)){
  dir.create(dirname(parcel_file), recursive = T)
  download.file(url = "https://openneuro.org/crn/datasets/ds000224/snapshots/00002/files/derivatives:surface_pipeline:sub-MSC01:surface_parcellation:sub-MSC01_parcels.dtseries.nii", destfile = parcel_file)
}

if(!file.exists(comm_file)){
  dir.create(dirname(comm_file), recursive = T)
  download.file(url = "https://openneuro.org/crn/datasets/ds000224/snapshots/00002/files/derivatives:surface_pipeline:sub-MSC01:surface_parcellation:sub-MSC01_parcel_networks.dscalar.nii", destfile = comm_file)
}

```

## Read in cifti files

* MSC-01's data are used
    + Individual specific parcellation and community (network) are loaded.
```{r, warning=F}
cii <- read_cifti(mscfile, drop_data = FALSE, trans_data = T) 

# Make brainstructure index
cii$brainstructureindex <- as.matrix(NA, dim(cii$data)[1])
for(i in 1:length(cii$BrainModel)){
  startindx <- as.numeric(attributes(cii$BrainModel[[i]])$IndexOffset + 1)
  endindx <- as.numeric(attributes(cii$BrainModel[[i]])$IndexOffset+attributes(cii$BrainModel[[i]])$IndexCount)
  
  cii$brainstructureindex[startindx:endindx] <- i
  
}

# Load parcels
parcel <- read_cifti(parcel_file)
parcel <- as.matrix(parcel$data)

# Load community
comm <- read_cifti(comm_file)
comm <- as.matrix(comm$data)
  
# Check dimension of cifti data (volume/frame x vertices)
dim(cii$data) # ~ 64k vertices, includes subcortical volumes
dim(parcel)   # surface only, excluded medial wall
dim(comm)     # surface only, excluded medial wall

# What are the labeled brain structures in the cii file? 
cifti_brain_structs(cii)

# ==== Take the Left & Right cortex only
cdata <- as.matrix(cii$data[cii$brainstructureindex==1 | cii$brainstructureindex==2,,])
# cdata <- cii$data[,,]
dim(cdata)

u_parcel <- unique(parcel)
u_parcel <- u_parcel[u_parcel!=0] # Remove parcel 0 and order parcel by #

```


## Extract Nodes' mean time series from surface data
```{r}
# ==== Mask out bad volumes from data
tmask <- read.table(tmask_file)$V1
ctmask <- cdata[,as.logical(tmask)]

# ==== Extract mean time series from each parcel -> Parcel x Vol matrix
tp <- matrix(0, length(u_parcel), sum(tmask))   

for(i in 1:length(u_parcel)){               
  tp[i,]<- colMeans(ctmask[which(parcel==u_parcel[i]),])
}

# order by parcel number
tp <- tp[order(u_parcel),]

```


## Plot processed mean time series of each node

* The heatmaps here are generated using a customized version of the [superheat (github)](https://github.com/mychan24/superheat) package. 

```{r tpmat}
superheat::superheat(tp,
                     heat.lim = c(-20, 20), 
                     heat.pal = c("black","white"),
                     grid.hline = FALSE,
                     grid.vline = FALSE,
                     title="Mean Time series of each parcel")
```

## Correlation Matrix (z-transformed)
```{r, fig.width=6, fig.height=7}
r <- cor(t(tp))         # Correlation matrix between all nodes
z <- psych::fisherz(r)  # Fisher's z-transform: 0.5 * log((1+r)/(1-r))

diag(z) <- 0            # Set diagonal to '0'; not informative

superheat::superheat(z, 
                     y.axis.reverse = TRUE, # Used to make origin (0,0) on top left corner
                     heat.lim = c(-.1, .5), 
                     heat.pal = parula(100),# rev(brewer.rdylbu(100)), 
                     heat.pal.values = c(0, 0.5,1),
                     grid.hline = FALSE,
                     grid.vline = FALSE,
                     title="Node x Node Correlation Matrix (z-transformed)")

```

## Correlation Matrix, nodes ordered by systems 

### Setup System Color for Plot
```{r}
parlabel <- data.frame(parcel_num=sort(u_parcel), community=NA)
plotlabel <- read.csv("../data/systemlabel_MSC.txt", header=F,
                          col.names = c("community","comm_label","color","comm_shortlabel"))

for(i in 1:length(u_parcel)){
  parlabel$community[i] <- unique(comm[which(parcel==sort(u_parcel)[i])])
}

```

```{r, fig.width=6, fig.height=7}
superheat::superheat(X = z, 
                     y.axis.reverse = TRUE,
                     membership.rows = parlabel$community,
                     membership.cols = parlabel$community,
                     left.label.col=plotlabel$color,
                     bottom.label.col=plotlabel$color,
                     extreme.values.na = FALSE,
                     heat.lim = c(-.2, .2), 
                     heat.pal = rev(brewer.rdylbu(100)),
                     heat.pal.values = c(0, 0.15, 0.25, 0.75,1),
                     title="Parcel x Parcel Correlation Matrix (z-transformed)")

```

## Splitting Negative and Positive 
```{r, fig.show='hide'}
# ==== Setup positive matrix plot
z_pos <- z
z_pos[z<0] <- 0
ss_pos <- superheat::superheat(X = z_pos, 
                     y.axis.reverse = TRUE,
                     membership.rows = parlabel$community,
                     membership.cols = parlabel$community,
                     left.label.col=plotlabel$color,
                     bottom.label.col=plotlabel$color,
                     extreme.values.na = FALSE,
                     heat.lim = c(0, .3), 
                     heat.pal = parula(20),
                     heat.pal.values = c(0, 0.5, 1),
                     title="Node x Node Positive Correlation Matrix (z-transformed")

# ==== Setup negative matrix plot
z_neg <- z
z_neg[z>0] <- 0
ss_neg <- superheat::superheat(X = z_neg, 
                     y.axis.reverse = TRUE,
                     membership.rows = parlabel$community,
                     membership.cols = parlabel$community,
                     left.label.col=plotlabel$color,
                     bottom.label.col=plotlabel$color,
                     extreme.values.na = FALSE,
                     heat.lim = c(-.3, 0), 
                     heat.pal = rev(parula(20)),
                     heat.pal.values = c(0, 0.5, 1),
                     title="Node x Node Negative Correlation Matrix (z-transformed")

```


```{r pn_matrices, fig.width=12, fig.height=7}
gridExtra::grid.arrange(ggplotify::as.grob(ss_pos$plot), ggplotify::as.grob(ss_neg$plot), 
                        nrow=1)

```


## Plot Positive Netowrk Graph (requires "igraph")
```{r}
## Threshold matrix to 4%
z4 <- z_pos
z4[z < quantile(z, 0.96)] <- 0
net <- graph.adjacency(adjmatrix = z4, mode = "undirected", diag = F, weighted = T)

parlabel$id <- 1:nrow(parlabel)
parlabel$color <- NA
u_comm <- unique(parlabel$community)
for(i in u_comm){
  parlabel$color[parlabel$community==i] <- as.character(plotlabel$color[plotlabel$community==i])
}

V(net)$id <- parlabel$id
V(net)$community <- parlabel$community
net <- simplify(net, remove.multiple = F, remove.loops = T) 

plot(net, layout=layout_with_fr, vertex.label=NA, vertex.size=5, 
     vertex.color=parlabel$color, alpha=.6)
```

